---
title: "Theory"
output: html_document
date: "2023-03-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE}
library('PWEXP')
library('RColorBrewer')
set.seed(1818)
```
# 1. Distribution
## 1.1 Piecewise Exponential Distribution

The hazard function, cumulative hazard function, PDF, CDF, quantile
function of an exponential distribution r.v. $t$ are: \begin{align*}
    h(t)&=\lambda\\
    H(t)&=\lambda t\\
    f(t)&=\lambda e^{-\lambda t}\\
    F(t)&=1-e^{-\lambda t}\\
    Q(p)&=\frac{-\log(1-p)}{\lambda}
\end{align*} The hazard function, cumulative hazard function, PDF,
survival function, quantile function of a piecewise exponential
distribution r.v. $t$ with breakpoints $d_i$ are: \begin{align*}
    h(t)&=\begin{cases}
        \lambda_1, & t<d_1\\
        \lambda_2, & d_1\le t<d_2\\
        \vdots\\
        \lambda_{r+1}, & t\ge d_r
    \end{cases}\\
    H(t)&=\begin{cases}
        \lambda_1 t, & t<d_1\\
        (\lambda_1-\lambda_2)d_1+\lambda_2t , & d_1\le t<d_2\\
        \vdots\\
        \left[\sum_{i=1}^r(\lambda_i-\lambda_{i+1})d_i\right]+\lambda_{r+1}t, & t\ge d_r
    \end{cases}\\
    f(t)=h(t)e^{-H(t)}&=\begin{cases}
        \lambda_1 e^{-\lambda_1 t}, & t<d_1\\
        \lambda_2 e^{(\lambda_2-\lambda_1)d_1-\lambda_2t} , & d_1\le t<d_2\\
        \vdots\\
        \lambda_{r+1}e^{\left[\sum_{i=1}^r(\lambda_{i+1}-\lambda_{i})d_i\right]-\lambda_{r+1}t}, & t\ge d_r
    \end{cases}\\
    S(t)=e^{-H(t)}&=\begin{cases}
        e^{-\lambda_1 t}, & t<d_1\\
        e^{(\lambda_2-\lambda_1)d_1-\lambda_2t} , & d_1\le t<d_2\\
        \vdots\\
        e^{\left[\sum_{i=1}^r(\lambda_{i+1}-\lambda_{i})d_i\right]-\lambda_{r+1}t}, & t\ge d_r
    \end{cases}\\\\
    Q(p)&=\begin{cases}
        \frac{-\log(1-p)}{\lambda_1} & p< 1-e^{-\lambda d_1}\\
        \frac{(\lambda_2-\lambda_1)d_1-\log(1-p)}{\lambda_2} & 1-e^{-\lambda d_1}\le p< 1-e^{(\lambda_2-
        \lambda_1)d_1-\lambda_2 d_2}\\
        \cdots\\
        \frac{\left[\sum_{i=1}^r (\lambda_{i+1}-\lambda_i)d_i\right]-\log(1-p)}{\lambda_{r+1}} & p\ge 1-e^{\left[\sum_{i=1}^r(\lambda_{i+1}-\lambda_{i})d_i\right]-\lambda_{r+1}d_r}\\
    \end{cases}
\end{align*}

The ``PWEXP`` package provides ``dpwexp()``, ``ppwexp()``, ``qpwexp()``, ``rpwexp()`` functions to the piecewise exponential distribution:
```{r fig.show="hold", out.width="50%", out.height="100%", fig.width=6, fig.height=5.8}
# Left Figure ------------------------------------------------------------
# use rpwexp function to generate piecewise exp samples with rate 2, 1, 3
r_sample <- rpwexp(50000, rate=c(2, 1, 3), breakpoint=c(0.3, 0.8))
hist(r_sample, freq=F, breaks=200, main="Density of Piecewsie Exp Dist", xlab='t', xlim=c(0, 1.2))

# piecewise exp density with rate 2, 1, 3 
t <- seq(0, 1.5, 0.01)
f2 <- dpwexp(t, rate=c(2, 1, 3), breakpoint=c(0.3, 0.8))
points(t, f2, col='red', pch=16)

# exp distribution can be a special case of piecewise exp distribution
f1 <- dpwexp(t, rate=2) 
lines(t, f1, lwd=2)
legend('topright', c('exp dist with rate 2','piecewise exp dist with rate 2, 1, 3','histogram of piecewise exp dist with rate 2, 1, 3'), col=c('black','red'), fill=c(NA, NA, 'grey'), border=c('white', 'white', 'black'), lty=c(1, NA, NA), pch=c(NA, 16, NA), lwd=2)

# Right Figure ------------------------------------------------------------
# CDF of piecewise exp with rate 2, 1, 3
F2 <- ppwexp(t, rate=c(2, 1, 3), breakpoint=c(0.3, 0.8), lower.tail=T)
plot(t, F2, type='l', col='red', lwd=2, main="CDF and Quantile Function of Piecewsie Exp Dist", xlim=c(0, 1.5), ylim=c(0, 1.5))

# CDF of exp dist is compatible with our package
F1 <- ppwexp(t, rate=2, lower.tail=T)
lines(t, F1, lwd=2)

# plot quantile functions of both distributions
lines(F1, qpwexp(F1, rate=2, lower.tail=T), lty=2, lwd=2)
lines(F2, qpwexp(F2, rate=c(2, 1, 3), breakpoint=c(0.3,0.8), lower.tail=T), col='red', lty=2, lwd=2)

abline(0, 1, col='grey')
legend('topleft', c('CDF of piecewise exp with rate 2, 1, 3', 'quantile function of piecewise exp with rate 2, 1, 3', 'CDF of exp with rate 2', 'quantile function of exp with rate 2'), col=c('red', 'red', 'black', 'black'), lty=c(1, 2, 1, 2), lwd=2)
```

## 1.2. Conditional Piecewise Exponential Distribution

The conditional survival function, CDF, PFD and quantile function of an
exponential distribution $t$ given $t>T$ is \begin{align*}
    S(t|t>T)&=\frac{S(t)}{S(T)}=e^{\lambda T-\lambda t}\\
    F(t|t>T)&=1-\frac{S(t)}{S(T)}=1-e^{\lambda T-\lambda t}\\
    f(t|t>T)&=\lambda e^{\lambda T-\lambda t}\\
    Q(p|t>T)&=\frac{\lambda T-\log(1-p)}{\lambda}
\end{align*}

The conditional survival function and CDF of a piecewise exponential
distribution $t$ given $t>T$ is \begin{align*}
    S(t|t>T)&=\frac{S(t)}{S(T)}, \text{ then plug in $S(t)$, $S(T)$}\\
    F(t|t>T)&=1-\frac{S(t)}{S(T)}, \text{ then plug in $S(t)$, $S(T)$}
\end{align*} The conditional quantile function of a piecewise
exponential distribution $t$ given $t>T$ is \begin{align*}
    Q(p|t>T)=\begin{cases}
        \frac{\lambda_1 T-\log(1-p)}{\lambda_{1}}, & 
        \hspace{-60pt} p <F(d_{1}|t>T), T <d_{1}\\
        \frac{\left[\sum_{i=1}^{k-1} (\lambda_{i+1}-\lambda_i)d_i\right]+\lambda_1 T-\log(1-p)}{\lambda_{k}}, & \\
        &\hspace{-60pt} F(d_{k-1}|t>T)\le p <F(d_{k}|t>T), T <d_{1}\\
        \frac{\left[\sum_{i=m}^{k-1} (\lambda_{i+1}-\lambda_i)d_i\right]+\lambda_m T-\log(1-p)}{\lambda_{k}}, & \\
        &\hspace{-60pt} F(d_{k-1}|t>T)\le p <F(d_{k}|t>T), d_{m-1}\le T <d_{m}
    \end{cases}
\end{align*}
The ``PWEXP`` package provides ``ppwexp_conditional()``, ``qpwexp_conditional()``, ``rpwexp_conditional()`` functions for conditional piecewise exponential distribution:
```{r fig.show="hold", out.width="50%", out.height="100%", fig.width=6, fig.height=5.8}
# Left Figure ------------------------------------------------------------
# CDF and qunatile function of conditional piecewise exp with rate 2, 1, 3 given t > 0.1
t <- seq(0.1, 1.2, 0.01)
F2_con <- ppwexp_conditional(t, qT=0.1, rate=c(2, 1, 3), breakpoint=c(0.3, 0.8))
plot(t, F2_con, type='l', col='red', lwd=2, main="CDF and Quantile Function of Conditional \nPiecewsie Exp Dist", xlim=c(0, 1.2), ylim=c(0, 1.2))
lines(F2_con, qpwexp_conditional(F2_con, qT=0.1, rate=c(2, 1, 3), breakpoint=c(0.3,0.8)), lty=2, lwd=2, col='red')

# compare with CDF and quantile function of unconditional piecewise exp with rate 2, 1, 3
t <- seq(0, 1.2, 0.01)
F2 <- ppwexp(t, rate=c(2, 1, 3), breakpoint=c(0.3,0.8))
lines(t, F2, lwd=2)
lines(F2, qpwexp(F2, rate=c(2, 1, 3), breakpoint=c(0.3,0.8)), lty=2, lwd=2)
abline(v=0.1, col='grey')
abline(h=0.1, col='grey')
legend('topleft', c('CDF of piecewise exp dist given t > 0.1', 'quantile function of piecewise exp dist given t > 0.1', 'CDF of piecewise exp dist', 'quantile function of piecewise exp dist'), col=c('red', 'red', 'black', 'black'), lty=c(1, 2, 1, 2), lwd=2)

# Right Figure ------------------------------------------------------------
# use rpwexp_conditional function to generate piecewise exp samples with rate 2, 1, 3 given t > 0.1
r_sample_con <- rpwexp_conditional(3000, qT=0.1, rate=c(2, 1, 3), breakpoint=c(0.3,0.8))
plot(ecdf(r_sample_con), col='red', lwd=2,  main="Empirical CDF of Conditional Piecewsie Exp Dist", xlim=c(0, 1.2), ylim=c(0, 1))

# compare with its CDF
lines(seq(0.1, 1.2, 0.01), F2_con, lwd=2)
legend('topleft', c('empirial CDF of piecewise exp dist given t > 0.1', 'true CDF of piecewise exp dist given t > 0.1'), col=c('red', 'black'), lty=c(1,2), lwd=2)
```
