---
title: "Piecewise Exponential Prediction Model"
output: 
  html_document:
    toc: true
    code_folding: show
    toc_float: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = TRUE,
                      fig.height = 5,
                      fig.width = 10)

library(knitr)

```
<br><br>

# 0. Install Package 'ZIPFA'
You need to install package [optimx](https://cran.r-project.org/web/packages/optimx/index.html), [trustOptim](https://cran.r-project.org/web/packages/trustOptim/index.html) before installing [ZIPFA](https://cran.rstudio.com/web/packages/ZIPFA/index.html).
```{r eval=FALSE}
install.packages('optimx')
install.packages('trustOptim')
install.packages('ZIPFA')
library('Matrix')
library('parallel')
library('doParallel')
library('foreach')
library('optimx')
library('trustOptim')
library('ZIPFA')
```
```{r echo=FALSE, message=FALSE}
library('Matrix')
```

<br><br>

# 1. Piecewise Exponential Distribution

The hazard function, cumulative hazard function, PDF, CDF, quantile function of an exponential distribution r.v. $t$ is:
\begin{align*}
    h(t)&=\lambda\\
    H(t)&=\lambda t\\
    f(t)&=\lambda e^{-\lambda t}\\
    F(t)&=1-e^{-\lambda t}\\
    Q(p)&=\frac{-\log(1-p)}{\lambda}
\end{align*}
The hazard function, cumulative hazard function, PDF, survival function, quantile function of a piecewise exponential distribution r.v. $t$ is:
\begin{align*}
    h(t)&=\begin{cases}
        \lambda_1, & t<d_1\\
        \lambda_2, & d_1\le t<d_2\\
        \vdots\\
        \lambda_{r+1}, & t\ge d_r
    \end{cases}\\
    H(t)&=\begin{cases}
        \lambda_1 t, & t<d_1\\
        (\lambda_1-\lambda_2)d_1+\lambda_2t , & d_1\le t<d_2\\
        \vdots\\
        \left[\sum_{i=1}^r(\lambda_i-\lambda_{i+1})d_i\right]+\lambda_{r+1}t, & t\ge d_r
    \end{cases}\\
    f(t)=h(t)e^{-H(t)}&=\begin{cases}
        \lambda_1 e^{-\lambda_1 t}, & t<d_1\\
        \lambda_2 e^{(\lambda_2-\lambda_1)d_1-\lambda_2t} , & d_1\le t<d_2\\
        \vdots\\
        \lambda_{r+1}e^{\left[\sum_{i=1}^r(\lambda_{i+1}-\lambda_{i})d_i\right]-\lambda_{r+1}t}, & t\ge d_r
    \end{cases}\\
    S(t)=e^{-H(t)}&=\begin{cases}
        e^{-\lambda_1 t}, & t<d_1\\
        e^{(\lambda_2-\lambda_1)d_1-\lambda_2t} , & d_1\le t<d_2\\
        \vdots\\
        e^{\left[\sum_{i=1}^r(\lambda_{i+1}-\lambda_{i})d_i\right]-\lambda_{r+1}t}, & t\ge d_r
    \end{cases}\\\\
    Q(p)&=\begin{cases}
        \frac{-\log(1-p)}{\lambda_1} & p< 1-e^{-\lambda d_1}\\
        \frac{(\lambda_2-\lambda_1)d_1-\log(1-p)}{\lambda_2} & 1-e^{-\lambda d_1}\le p< 1-e^{(\lambda_2-
        \lambda_1)d_1-\lambda_2 d_2}\\
        \cdots\\
        \frac{\left[\sum_{i=1}^r (\lambda_{i+1}-\lambda_i)d_i\right]-\log(1-p)}{\lambda_{r+1}} & p\ge 1-e^{\left[\sum_{i=1}^r(\lambda_{i+1}-\lambda_{i})d_i\right]-\lambda_{r+1}d_r}\\
    \end{cases}
\end{align*}


# 2. Conditional Piecewise Exponential Distribution
The conditional survival function, CDF, PFD and quantile function of an exponential distribution $t$ given $t>T$ is 
\begin{align*}
    S(t|t>T)&=\frac{S(t)}{S(T)}=e^{\lambda T-\lambda t}\\
    F(t|t>T)&=1-\frac{S(t)}{S(T)}=1-e^{\lambda T-\lambda t}\\
    f(t|t>T)&=\lambda e^{\lambda T-\lambda t}\\
    Q(p|t>T)&=\frac{\lambda T-\log(1-p)}{\lambda}
\end{align*}

The conditional survival function and CDF of a piecewise exponential distribution $t$ given $t>T$ is 
\begin{align*}
    S(t|t>T)&=\frac{S(t)}{S(T)}, \text{ then plug in $S(t)$, $S(T)$}\\
    F(t|t>T)&=1-\frac{S(t)}{S(T)}, \text{ then plug in $S(t)$, $S(T)$}
\end{align*}
The conditional quantile function of a piecewise exponential distribution $t$ given $t>T$ is 
\begin{align*}
    Q(p|t>T)=\begin{cases}
        \frac{\lambda_1 T-\log(1-p)}{\lambda_{1}}, & 
        \hspace{-60pt} p <F(d_{1}|t>T), T <d_{1}\\
        \frac{\left[\sum_{i=1}^{k-1} (\lambda_{i+1}-\lambda_i)d_i\right]+\lambda_1 T-\log(1-p)}{\lambda_{k}}, & \\
        &\hspace{-60pt} F(d_{k-1}|t>T)\le p <F(d_{k}|t>T), T <d_{1}\\
        \frac{\left[\sum_{i=m}^{k-1} (\lambda_{i+1}-\lambda_i)d_i\right]+\lambda_m T-\log(1-p)}{\lambda_{k}}, & \\
        &\hspace{-60pt} F(d_{k-1}|t>T)\le p <F(d_{k}|t>T), d_{m-1}\le T <d_{m}
    \end{cases}
\end{align*}


